---
title: "Coronavirus in the Wisconsin: How Fast It's Growing"
# author: "Srikanth Aravamuthan & Sean Kent"
date: "Updated `r format(Sys.time(), '%B %d, %Y, %I:00 %p %Z')`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  # cache = TRUE, 
  message = FALSE,
  warning = FALSE, 
  fig.width = 10
)

library(tidyverse)
library(purrr)
library(lubridate)
library(zoo)
library(readxl)
library(maps)
# library(sf)
library(plotly)
library(shiny)

```



```{r data choices}
## Data choices
MIN_CASES_TO_PLOT <- 30
MIN_DATE <- Sys.Date() - 28 # last 4 weeks
MAX_DATE <- Sys.Date()

color_palette <- c(RColorBrewer::brewer.pal(21, "Paired"), rep("grey50", 20))
```



```{r nyt-data}
# Data from the New York Times repository
us_counties <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
us_states <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

# NYT does this annoying thing where they count ("Kings", "Queens", "New York", "Bronx") counties 
# as the 'county' New York City.  Attempt to fix by removing putting all of the cases/deaths in 
# New York county, since we are aggregating by metro area anyways
us_counties[us_counties$county == "New York City", "county"] <- "New York"

```



```{r population-data}
## Population data from US census 2019 estimates
county_population <- 
  read_csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/totals/co-est2019-alldata.csv") %>% 
  filter(SUMLEV == "050") %>% 
  mutate(CTYNAME = str_remove(CTYNAME, " County"),
         CTYNAME = str_remove(CTYNAME, " Parish")) %>% 
  rename(state = STNAME,
         county = CTYNAME,
         population = POPESTIMATE2019) %>% 
  select(state, county, population)

```



```{r}
gundersen <- 
  us_counties %>% 
  filter(county == "Adams" & state == "Wisconsin" |
         county == "Buffalo" & state == "Wisconsin" |
         county == "Crawford" & state == "Wisconsin" |
         county == "Grant" & state == "Wisconsin" |
         county == "Jackson" & state == "Wisconsin" |
         county == "Juneau" & state == "Wisconsin" |
         county == "La Crosse" & state == "Wisconsin" |
         county == "Marquette" & state == "Wisconsin" |
         county == "Monroe" & state == "Wisconsin" |
         county == "Richland" & state == "Wisconsin" |
         county == "Trempealeau" & state == "Wisconsin" |
         county == "Vernon" & state == "Wisconsin" |
         county == "Fillmore" & state == "Minnesota" |
         county == "Houston" & state == "Minnesota" |
         county == "Wabasha" & state == "Minnesota" |
         county == "Winona" & state == "Minnesota" |
         county == "Allamakee" & state == "Iowa" |
         county == "Clayton" & state == "Iowa" |
         county == "Fayette" & state == "Iowa" |
         county == "Howard" & state == "Iowa" |
         county == "Winneshiek" & state == "Iowa")
  
```

## Daily Growth Rate

#### Daily growth rate of confirmed cases

```{r dpi = 96}
plot_data <- 
  gundersen %>% 
  mutate(county.state = paste0(county,", ",state)) %>% 
  group_by(county.state) %>% 
  arrange(date) %>% 
  mutate(daily.change = (cases / lag(cases, n = 7, default = NA))^(1 / 7) - 1 ) %>% 
  ungroup() %>% 
  mutate(`Average Daily Change (Last 7 Days)` = daily.change) %>% 
  rename(Name = county.state,
         Date = date,
         Cases = cases) %>% 
  filter(between(Date, MIN_DATE, MAX_DATE))

fig1 <- plot_ly(data = plot_data,
               x = ~Date,
               y = ~`Average Daily Change (Last 7 Days)`,
               color = ~Name,
               text = ~paste0("</br>", Name,
                             "</br>", Date,
                             "</br>Daily Change: ", scales::percent(daily.change, accuracy = 0.1)),
               type = "scatter",
               mode = 'lines+markers',
               colors = "Paired",
               alpha = 0.7,
               hoverinfo = "text") %>% 
  layout(title = "") %>% 
  layout(xaxis = list(title = "Date"),
         yaxis = list(title = "Average Daily Change (Last 7 Days)",
                      range = c(0, 1),
                      tickformat = "%")) 

fig1

```

## Growth Rates by Case Count

#### Daily growth rate of confirmed cases

```{r dpi = 96}
plot_data <- 
  gundersen %>% 
  left_join(county_population, by = c("state", "county")) %>% 
  mutate(county.state = paste0(county,", ",state)) %>% 
  group_by(county.state) %>% 
  arrange(date) %>% 
  mutate(daily.change = (cases / lag(cases, n = 7, default = NA))^(1 / 7) - 1 ) %>% 
  ungroup() %>% 
  mutate(`Confirmed Cases Per Thousand People` = cases / population * 1000) %>% 
  mutate(`Average Daily Change (Last 7 Days)` = daily.change) %>% 
  rename(Name = county.state,
         Date = date,
         Cases = cases) %>% 
  filter(between(Date, MIN_DATE, MAX_DATE))

fig2 <- plot_ly(data = plot_data,
               x = ~`Confirmed Cases Per Thousand People`,
               y = ~`Average Daily Change (Last 7 Days)`,
               color = ~Name,
               text = ~paste0("</br>", Name,
                             "</br>", Date,
                             "</br>Cases Per Thousand: ", 
                             round(`Confirmed Cases Per Thousand People`, 3), 
                             "</br>Daily Change: ", scales::percent(daily.change, accuracy = 0.1)),
               type = "scatter",
               mode = 'lines+markers',
               colors = "Paired",
               # alpha = 0.5,
               hoverinfo = "text") %>% 
  layout(title = "") %>% 
  layout(xaxis = list(title = "Confirmed Cases Per Thousand People"),
         yaxis = list(title = "Average Daily Change (Last 7 Days)",
                      range = c(0, 1),
                      tickformat = "%")) 

fig2

```

## New Cases by Case Count

#### Weekly new cases

```{r dpi = 96}
plot_data <- 
  gundersen %>% 
  left_join(county_population, by = c("state", "county")) %>% 
  mutate(county.state = paste0(county,", ",state)) %>% 
  group_by(county.state) %>% 
  arrange(date) %>% 
  mutate(new.cases = cases - lag(cases, 
                                 default = cases[1]),
         roll.cases = rollapply(new.cases, 
                                list(-(0:6)), 
                                sum, 
                                fill = NA, 
                                align = "right", 
                                partial = FALSE)) %>% 
  ungroup() %>% 
  mutate(`New Cases (Daily Average over Last 7 Days)` = roll.cases) %>% 
  rename(Name = county.state,
         Date = date,
         Cases = cases) %>% 
  filter(between(Date, MIN_DATE, MAX_DATE))

fig3 <- plot_ly(data = plot_data,
               x = ~Cases,
               y = ~`New Cases (Daily Average over Last 7 Days)`,
               color = ~Name,
               text = ~paste0("</br>", Name,
                             "</br>", Date,
                             "</br>Cases: ", Cases, 
                             "</br>New Cases (Daily Average over Last 7 Days): ", `New Cases (Daily Average over Last 7 Days)`),
               type = "scatter",
               mode = 'lines+markers',
               colors = "Paired",
               # alpha = 0.5,
               hoverinfo = "text") %>% 
  layout(yaxis = list(title = "New Cases (Daily Average over Last 7 Days)"))

fig3

```

## New Cases per Capita by Case Count per Capita

#### Weekly new cases

```{r dpi = 96}
plot_data <- 
  gundersen %>% 
  left_join(county_population, by = c("state", "county")) %>% 
  mutate(county.state = paste0(county,", ",state)) %>% 
  group_by(county.state) %>% 
  arrange(date) %>% 
  mutate(new.cases = cases - lag(cases, 
                                 default = cases[1]),
         roll.cases = rollapply(new.cases, 
                                list(-(0:6)), 
                                sum, 
                                fill = NA, 
                                align = "right", 
                                partial = FALSE)) %>% 
  ungroup() %>% 
  mutate(`New Cases Per Thousand People (Daily Average over Last 7 Days)` = round(roll.cases / population * 1000, 3)) %>% 
  mutate(`Cases Per Thousand People` = round(cases / population * 1000, 3)) %>% 
  rename(Name = county.state,
         Date = date,
         Cases = cases) %>% 
  filter(between(Date, MIN_DATE, MAX_DATE))

fig4 <- plot_ly(data = plot_data,
               x = ~`Cases Per Thousand People`,
               y = ~`New Cases Per Thousand People (Daily Average over Last 7 Days)`,
               color = ~Name,
               text = ~paste0("</br>", Name,
                             "</br>", Date,
                             "</br>Cases Per Thousand: ", `Cases Per Thousand People`, 
                             "</br>New Cases Per Thousand </br>(Daily Average over Last 7 Days): ", `New Cases Per Thousand People (Daily Average over Last 7 Days)`),
               type = "scatter",
               mode = 'lines+markers',
               colors = "Paired",
               # alpha = 0.5,
               hoverinfo = "text") %>% 
  layout(xaxis = list(title = "Cases Per Thousand People"),
         yaxis = list(title = "New Cases Per Thousand People (Daily Average over Last 7 Days)"))

fig4

```

<!-- ## Concept Plots -->

<!-- ```{r dpi = 96} -->
<!-- plot_data <-  -->
<!--   gundersen %>%  -->
<!--   left_join(county_population, by = c("state", "county")) %>%  -->
<!--   mutate(county.state = paste0(county,", ",state)) %>%  -->
<!--   group_by(county.state) %>%  -->
<!--   arrange(date) %>%  -->
<!--   mutate(new.cases = cases - lag(cases,  -->
<!--                                  default = cases[1]), -->
<!--          roll.cases = rollapply(new.cases,  -->
<!--                                 list(-(0:6)),  -->
<!--                                 sum,  -->
<!--                                 fill = NA,  -->
<!--                                 align = "right",  -->
<!--                                 partial = FALSE)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(`New Cases Per Thousand People (Daily Average over Last 7 Days)` = round(roll.cases / population * 1000, 3)) %>%  -->
<!--   rename(Name = county.state, -->
<!--          Date = date, -->
<!--          Cases = cases) %>%  -->
<!--   filter(between(Date, MIN_DATE, MAX_DATE)) -->

<!-- fig5 <- plot_ly(data = plot_data, -->
<!--                x = ~Cases, -->
<!--                y = ~`New Cases Per Thousand People (Daily Average over Last 7 Days)`, -->
<!--                color = ~Name, -->
<!--                text = ~paste0("</br>", Name, -->
<!--                              "</br>", Date, -->
<!--                              "</br>Cases: ", Cases,  -->
<!--                              "</br>New Cases Per Thousand </br>(Daily Average over Last 7 Days): ", `New Cases Per Thousand People (Daily Average over Last 7 Days)`), -->
<!--                type = "scatter", -->
<!--                mode = 'lines+markers', -->
<!--                colors = "Paired", -->
<!--                # alpha = 0.5, -->
<!--                hoverinfo = "text") %>%  -->
<!--   layout(title = "") -->

<!-- fig5 -->

<!-- ``` -->

<!-- ```{r dpi = 96} -->
<!-- plot_data <-  -->
<!--   gundersen %>%  -->
<!--   left_join(county_population, by = c("state", "county")) %>%  -->
<!--   mutate(county.state = paste0(county,", ",state)) %>%  -->
<!--   group_by(county.state) %>%  -->
<!--   arrange(date) %>%  -->
<!--   mutate(new.cases = cases - lag(cases,  -->
<!--                                  default = cases[1]), -->
<!--          roll.cases = rollapply(new.cases,  -->
<!--                                 list(-(0:6)),  -->
<!--                                 sum,  -->
<!--                                 fill = NA,  -->
<!--                                 align = "right",  -->
<!--                                 partial = FALSE)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(`New Cases (Daily Average over Last 7 Days)` = roll.cases) %>%  -->
<!--   rename(Name = county.state, -->
<!--          Date = date, -->
<!--          Cases = cases) %>%  -->
<!--   filter(between(Date, MIN_DATE, MAX_DATE)) -->

<!-- fig6 <- plot_ly(data = plot_data, -->
<!--                x = ~Date, -->
<!--                y = ~`New Cases (Daily Average over Last 7 Days)`, -->
<!--                color = ~Name, -->
<!--                text = ~paste0("</br>", Name, -->
<!--                              "</br>", Date, -->
<!--                              "</br>Cases: ", Cases,  -->
<!--                              "</br>New Cases (Daily Average over Last 7 Days): ", `New Cases (Daily Average over Last 7 Days)`), -->
<!--                type = "scatter", -->
<!--                mode = 'lines+markers', -->
<!--                colors = "Paired", -->
<!--                # alpha = 0.5, -->
<!--                hoverinfo = "text") %>%  -->
<!--   layout(title = "") -->

<!-- fig6 -->

<!-- ``` -->

<!-- ```{r dpi = 96} -->
<!-- plot_data <-  -->
<!--   gundersen %>%  -->
<!--   left_join(county_population, by = c("state", "county")) %>%  -->
<!--   mutate(county.state = paste0(county,", ",state)) %>%  -->
<!--   group_by(county.state) %>%  -->
<!--   arrange(date) %>%  -->
<!--   mutate(new.cases = cases - lag(cases,  -->
<!--                                  default = cases[1]), -->
<!--          roll.cases = rollapply(new.cases,  -->
<!--                                 list(-(0:6)),  -->
<!--                                 sum,  -->
<!--                                 fill = NA,  -->
<!--                                 align = "right",  -->
<!--                                 partial = FALSE)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(`New Cases Per Thousand People (Daily Average over Last 7 Days)` = round(roll.cases / population * 1000, 3)) %>%  -->
<!--   rename(Name = county.state, -->
<!--          Date = date, -->
<!--          Cases = cases) %>%  -->
<!--   filter(between(Date, MIN_DATE, MAX_DATE)) -->

<!-- fig7 <- plot_ly(data = plot_data, -->
<!--                x = ~Date, -->
<!--                y = ~`New Cases Per Thousand People (Daily Average over Last 7 Days)`, -->
<!--                color = ~Name, -->
<!--                text = ~paste0("</br>", Name, -->
<!--                              "</br>", Date, -->
<!--                              "</br>Cases: ", Cases,  -->
<!--                              "</br>New Cases Per Thousand </br>(Daily Average over Last 7 Days): ", `New Cases Per Thousand People (Daily Average over Last 7 Days)`), -->
<!--                type = "scatter", -->
<!--                mode = 'lines+markers', -->
<!--                colors = "Paired", -->
<!--                # alpha = 0.5, -->
<!--                hoverinfo = "text") %>%  -->
<!--   layout(title = "") -->

<!-- fig7 -->

<!-- ``` -->

<!-- ```{r dpi = 96} -->
<!-- plot_data <-  -->
<!--   gundersen %>%  -->
<!--   left_join(county_population, by = c("state", "county")) %>%  -->
<!--   mutate(county.state = paste0(county,", ",state)) %>%  -->
<!--   group_by(county.state) %>%  -->
<!--   arrange(date) %>%  -->
<!--   mutate(new.cases = cases - lag(cases,  -->
<!--                                  default = cases[1]), -->
<!--          roll.cases = rollapply(new.cases,  -->
<!--                                 list(-(0:6)),  -->
<!--                                 sum,  -->
<!--                                 fill = NA,  -->
<!--                                 align = "right",  -->
<!--                                 partial = FALSE)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(`New Cases (Daily Average over Last 7 Days)` = roll.cases) %>%  -->
<!--   mutate(`Cases Per Thousand People` = round(cases / population * 1000, 3)) %>%  -->
<!--   rename(Name = county.state, -->
<!--          Date = date, -->
<!--          Cases = cases) %>%  -->
<!--   filter(between(Date, MIN_DATE, MAX_DATE)) -->

<!-- fig8 <- plot_ly(data = plot_data, -->
<!--                x = ~`Cases Per Thousand People`, -->
<!--                y = ~`New Cases (Daily Average over Last 7 Days)`, -->
<!--                color = ~Name, -->
<!--                text = ~paste0("</br>", Name, -->
<!--                              "</br>", Date, -->
<!--                              "</br>Cases Per Thousand: ", `Cases Per Thousand People`,  -->
<!--                              "</br>New Cases (Daily Average over Last 7 Days): ", `New Cases (Daily Average over Last 7 Days)`), -->
<!--                type = "scatter", -->
<!--                mode = 'lines+markers', -->
<!--                colors = "Paired", -->
<!--                # alpha = 0.5, -->
<!--                hoverinfo = "text") %>%  -->
<!--   layout(title = "") -->

<!-- fig8 -->

<!-- ``` -->

## Map of Gundersen Clinic Areas and Adjacent Counties

#### Confirmed cases per thousand residents

```{r}
county <-
  maps::map("county", plot = FALSE, fill = TRUE) %>%
  sf::st_as_sf() %>%
  separate(ID, c("state", "county"), ",") %>%
  filter(state %in% c("wisconsin","minnesota","iowa","illinois")) %>% 
  mutate(state = tools::toTitleCase(state),
         county = tools::toTitleCase(county))

county_adjacency <- 
  read_tsv("https://www2.census.gov/geo/docs/reference/county_adjacency.txt",
               col_names = FALSE) %>% 
  fill(X1:X2) %>% 
  separate(X1, c("county", "state"), ", ") %>% 
  separate(X3, c("county.adjacent", "state.adjacent"), ", ") %>% 
  rename(fips = X2,
         fips.adjacent = X4)

county_adjacency <- 
  county_adjacency %>% 
  filter(county == "Adams County" & state == "WI" |
         county == "Buffalo County" & state == "WI" |
         county == "Crawford County" & state == "WI" |
         county == "Grant County" & state == "WI" |
         county == "Jackson County" & state == "WI" |
         county == "Juneau County" & state == "WI" |
         county == "La Crosse County" & state == "WI" |
         county == "Marquette County" & state == "WI" |
         county == "Monroe County" & state == "WI" |
         county == "Richland County" & state == "WI" |
         county == "Trempealeau County" & state == "WI" |
         county == "Vernon County" & state == "WI" |
         county == "Fillmore County" & state == "MN" |
         county == "Houston County" & state == "MN" |
         county == "Wabasha County" & state == "MN" |
         county == "Winona County" & state == "MN" |
         county == "Allamakee County" & state == "IA" |
         county == "Clayton County" & state == "IA" |
         county == "Fayette County" & state == "IA" |
         county == "Howard County" & state == "IA" |
         county == "Winneshiek County" & state == "IA") %>% 
  select(county.adjacent, state.adjacent) %>% 
  rename(county = county.adjacent,
         state = state.adjacent) %>% 
  mutate(county = str_remove(county, " County"),
         state = case_when(
           state == "WI" ~ "Wisconsin", 
           state == "MN" ~ "Minnesota", 
           state == "IA" ~ "Iowa", 
           state == "IL"  ~ "Illinois")
         ) %>% 
  distinct(county, state)

gundersen_current_date <- 
  us_counties %>% 
  right_join(county_adjacency, by = c("county", "state")) %>% 
  filter(date == max(date, na.rm = TRUE) | is.na(date)) %>% 
  mutate(clinic.area = TRUE)

```

```{r dpi = 96}
## US Metro Map (Cases)
fig9 <- county %>%
  left_join(gundersen_current_date, by = c("state", "county")) %>%
  left_join(county_population, by = c("state", "county")) %>% 
  # filter(!is.na(clinic.area)) %>%
  mutate(name = ifelse(is.na(clinic.area), state, paste0(county, ", ", state))) %>% 
  group_by(name) %>%
  summarize(cases = sum(cases, na.rm = TRUE),
            deaths = sum(deaths, na.rm = TRUE),
            population = max(population, na.rm = TRUE),
            geometry = sf::st_union(sf::st_buffer(geometry,0.0))) %>%
  ungroup() %>%
  mutate(cases.per.1000 = round(cases / (population / 1000), 3),
         # cases.per.1000 = ifelse(cases.per.1000<=0.025, NA, cases.per.1000),
         cases = ifelse(name %in% c("Wisconsin","Minnesota","Iowa","Illinois"), NA, cases),
         population = ifelse(name %in% c("Wisconsin","Minnesota","Iowa","Illinois"), NA, population),
         cases.per.1000 = ifelse(name %in% c("Wisconsin","Minnesota","Iowa","Illinois"), NA, cases.per.1000),
         name = ifelse(name %in% c("Wisconsin","Minnesota","Iowa","Illinois"), paste0("Rest of ", name), name)) %>% 
  mutate(cases.per.1000.tool.tip = cases.per.1000,
         cases.per.1000.rank = rank(-cases.per.1000, na.last = "keep", ties.method = "min"),
         cases.per.1000 = cut(cases.per.1000,
                              c(min(cases.per.1000, na.rm = TRUE), 0.1, 0.25, 0.5, 1, 2, 4, max(cases.per.1000, na.rm = TRUE)))) %>%
  # sf::st_transform(crs = "+proj=aea +lat_1=25 +lat_2=50 +lon_0=-100") %>% 
  sf::st_cast("MULTIPOLYGON") %>%
  plot_ly(split = ~name, 
          color = ~cases.per.1000, 
          colors = "YlGnBu", 
          span = I(1),
          stroke = I("gray50"),
          alpha = 1,
          text = ~paste0("</br>", name, 
                         "</br>Cases: ", cases, 
                         "</br>Population: ", population, 
                         "</br>Cases per Thousand: ", cases.per.1000.tool.tip,
                         " (", cases.per.1000.rank, " out of ", nrow(.)-4, ")"),
          hoverinfo = "text",
          hoveron = "fills") %>% 
  layout(title="",
         showlegend = FALSE) 
# %>%
#   colorbar(title = "Cases per Thousand")

fig9

```



## Sources

New York Times. New York Times database of U.S. coronavirus cases. 2020. Retrieved from <https://github.com/nytimes/covid-19-data>.

United States Census Bureau, County Population Totals: 2010-2019. 2019. Retrieved from <https://www.census.gov/data/tables/time-series/demo/popest/2010s-counties-total.html>.

United States Census Bureau, County Adjacency File. 2010. Retrieved from <https://www.census.gov/geographies/reference-files/2010/geo/county-adjacency.html>.

## Appendix: Calculating Average Daily Change

Calculate "Average Daily Change in Total Cases, Over the Previous 7 Days" at time $t$ as

$$
\textrm{avg daily change}_t =  \left( \frac{\textrm{cases}_t}{\textrm{cases}_{t-7}} \right)^{(1/7)} - 1 
$$

## Appendix: Calculating New Cases (Daily Average over Last 7 Days)

Calculate "New Cases (Daily Average over Last 7 Days)" at time $t$ as

$$
\textrm{new cases}_t =  \frac{1}{7}\left( \textrm{cases}_t-\textrm{cases}_{t-7}\right)
$$



<!-- ## Appendix: R Code -->

```{r appendix, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
