---
title: "Coronavirus Burden and Intensity Across Wisconsin"
# author: "Srikanth Aravamuthan & Sean Kent"
date: "Updated `r format(Sys.time(), '%B %d, %Y, %I:00 %p %Z')`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  # cache = TRUE, 
  message = FALSE,
  warning = FALSE, 
  fig.width = 10
)

library(tidyverse)
library(purrr)
library(lubridate)
library(zoo)
library(readxl)
library(here)
# library(maps)
# library(sf)
library(plotly)
# library(shiny)

# library(maptools)
# library(rgeos)

```



```{r data choices}
## Data choices
MIN_CASES_TO_PLOT <- 30
MIN_DATE <- Sys.Date() - 28 # last 4 weeks
MAX_DATE <- Sys.Date()

# color_palette <- c(RColorBrewer::brewer.pal(21, "Paired"), rep("grey50", 20))
```



```{r nyt-data}
# Data from the New York Times repository
us_counties <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
us_states <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

# NYT does this annoying thing where they count ("Kings", "Queens", "New York", "Bronx") counties 
# as the 'county' New York City.  Attempt to fix by removing putting all of the cases/deaths in 
# New York county, since we are aggregating by metro area anyways
us_counties[us_counties$county == "New York City", "county"] <- "New York"

```



```{r population-data}
## Population data from US census 2019 estimates
county_population <- 
  read_csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/totals/co-est2019-alldata.csv") %>% 
  filter(SUMLEV == "050") %>% 
  mutate(CTYNAME = str_remove(CTYNAME, " County"),
         CTYNAME = str_remove(CTYNAME, " Parish")) %>% 
  rename(state = STNAME,
         county = CTYNAME,
         population = POPESTIMATE2019) %>% 
  select(state, county, population)

```



```{r}

wi_dph_regions  <- 
  read_csv(here("DHS/wi_dph_regions.csv")) %>% 
  select(county, dph_region,  herc_region)


wi_counties <- 
  us_counties %>% 
  filter(county %in% wi_dph_regions$county & state == "Wisconsin") %>% 
  left_join(wi_dph_regions, by = "county") %>% 
  left_join(county_population, by = c("county", "state"))

```


```{r}
wi_regions <- 
  wi_counties %>% 
  group_by(date, herc_region) %>% 
  summarize(state = unique(state),
            cases = sum(cases),
            deaths = sum(deaths),
            population = sum(population)) %>% 
  ungroup()
```



The following is a *very* rough draft of visualizations to track the burden and intensity of HERC regions in Wisconsin. 


## Burden vs Intensity


```{r}

process_data <- function(data, group_var) {
  group_var <- enquo(group_var)
  
  data %>% 
    ungroup() %>% 
    group_by(!! group_var) %>% 
    arrange(desc(date)) %>% 
    mutate(cases = cummin(cases)) %>% 
    arrange(date) %>% 
    mutate(new.cases = cases - lag(cases, default = 0)) %>% 
    mutate(cases.per.7days = rollapply(new.cases, 
                                       list(-(0:6)), 
                                       sum, 
                                       fill = NA,
                                       align = "right", 
                                       partial = FALSE)) %>% 
    mutate(cases.per.14days = rollapply(new.cases, 
                                        list(-(0:13)), 
                                        sum, 
                                        fill = NA,
                                        align = "right", 
                                        partial = FALSE)) %>% 
    mutate(cases.per.14days.per.100k = cases.per.14days / (population/100000)) %>% 
    mutate(row = row_number()) %>% 
    filter(between(date, MIN_DATE, MAX_DATE))
    
  # TODO: add select to narrow down the variables kept
}

# process_data(wi_regions, group_var = herc_region)
# process_data(wi_counties, group_var = county) 

```



### HERC Regions Across Wisconsin (Last 4 weeks)


```{r}
fig1 <- 
    plot_ly(data = process_data(wi_regions, group_var = herc_region),
            x = ~cases.per.14days.per.100k,
            y = ~cases.per.7days,
            color = ~herc_region,
            # size = ~row, 
            # color = I("black"),
            colors = "Dark2", 
            text = ~paste0("</br>", herc_region,
                           "</br>", date,
                           "</br>New Cases (last 7 days): ", cases.per.7days,
                           "</br>New Cases per 100k (last 14 days): ", round(cases.per.14days.per.100k, 1)),
            type = "scatter",
            mode = 'lines+markers',
            hoverinfo = "text") %>% 
    layout(title = list(text = "",
                        y = 1,
                        x = 0.1,
                        xanchor = "left")) %>% 
    layout(xaxis = list(title = "Burden: New Cases per 100k (last 14 days)"),
           yaxis = list(title = "Intensity: New Cases (last 7 days)")) %>% 
    config(doubleClickDelay = 500)
  
fig1
  
```


### Individual Regions (Last 4 Weeks)

```{r}
# region = "South Central"
l <- htmltools::tagList()

regions <- unique(wi_dph_regions$herc_region)
for (i in 1:length(regions)) {
  fig2 <- 
    plot_ly(data = process_data(wi_regions, group_var = herc_region) %>% 
              filter(herc_region == regions[i]),
            x = ~cases.per.14days.per.100k,
            y = ~cases.per.7days,
            # size = ~row, 
            name = regions[i],
            color = I("black"),
            text = ~paste0("</br>", herc_region,
                           "</br>", date,
                           "</br>New Cases (last 7 days): ", cases.per.7days,
                           "</br>New Cases per 100k (last 14 days): ", round(cases.per.14days.per.100k, 1)),
            type = "scatter",
            mode = 'lines+markers',
            hoverinfo = "text") 
  
  fig2 <- add_trace(fig2, 
                    data = process_data(wi_counties, group_var = county) %>% 
                      filter(herc_region == regions[i]),
                    inherit = FALSE,
                    x = ~cases.per.14days.per.100k,
                    y = ~cases.per.7days,
                    color = ~county,
                    text = ~paste0("</br>", county,
                                   "</br>", date,
                                   "</br>New Cases (last 7 days): ", cases.per.7days,
                                   "</br>New Cases per 100k (last 14 days): ", round(cases.per.14days.per.100k, 1)),
                    type = "scatter",
                    mode = 'lines+markers',
                    colors = "Paired",
                    alpha = 0.7,
                    hoverinfo = "text"
  )
  
  # formatting 
  fig2 <- fig2 %>% 
    layout(title = list(text = paste0(regions[i], " Region"),
                        y = 1,
                        x = 0.1,
                        xanchor = "left")) %>% 
    layout(xaxis = list(title = "Burden: New Cases per 100k (last 14 days)"),
           yaxis = list(title = "Intensity: New Cases (last 7 days)")) %>% 
    config(doubleClickDelay = 500)
  
  l[[i]] <- as_widget(fig2)
  
}

l


```


*These plots are interactive*: Double click on a region in the legend to start a comparison and then single click on others to add or remove them from the comparison.  Note: double-clicking may not work well on all platforms, but you may single click on a region in the legend to remove from the comparison.  




## Sources

New York Times. New York Times database of U.S. coronavirus cases. 2020. Retrieved from <https://github.com/nytimes/covid-19-data>.

United States Census Bureau, County Population Totals: 2010-2019. 2019. Retrieved from <https://www.census.gov/data/tables/time-series/demo/popest/2010s-counties-total.html>.

United States Census Bureau, County Adjacency File. 2010. Retrieved from <https://www.census.gov/geographies/reference-files/2010/geo/county-adjacency.html>.



<!-- ## Appendix: R Code -->

```{r appendix, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
